---
title: "Deep Neural Network in R"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load library

```{r message=FALSE}
library(keras)
library(mlbench)
library(dplyr)
library(magrittr)
library(neuralnet)
```

## Getting data

```{r}
data("BostonHousing")
data <- BostonHousing
str(data)
```


This data set contains 506 observations and 14 variables. The variables chas as a factor variable, we need to convert in into numerical variable because neural network handles only numerical data.

Medv is the response variable, and the remaining are the predictors.

We need to convert factor variables into numeric variables while using the below-mentioned command. This function automatically detects the factor variables and convert them into numerical variables.

```{r}
data %<>% mutate_if(is.factor, as.numeric)
```


## Neural network visualization in R

Each predictor variable has one neuron, the first layer has 12 neurons, the second layer has 7 neurons and the out variable has one neuron.

```{r}
n <- neuralnet(medv ~ crim+zn+indus+chas+nox+rm+age+dis+rad+tax+ptratio+b+lstat,
               data = data,
               hidden = c(12,7),
               linear.output = F,
               lifesign = 'full',
               rep=1)
plot(n,col.hidden = 'darkgreen',     
col.hidden.synapse = 'darkgreen',
     show.weights = F,
     information = F,
     fill = 'lightblue')
```

## Data partition

```{r}
data <- as.matrix(data)
dimnames(data) <- NULL
set.seed(123)
ind <- sample(2, nrow(data), replace = T, prob = c(.7, .3))
training <- data[ind==1,1:13]
test <- data[ind==2, 1:13]
trainingtarget <- data[ind==1, 14]
testtarget <- data[ind==2, 14]
str(trainingtarget)
str(testtarget)
```


## Scaling features
```{r}
m <- colMeans(training)
s <- apply(training, 2, sd)
training <- scale(training, center = m, scale = s)
test <- scale(test, center = m, scale = s)
```

## Running model
```{r}
model <- keras_model_sequential()
# model %>%
#          layer_dense(units = 5, activation = 'relu', input_shape = c(13)) %>%
#          layer_dense(units = 1)
model %>%
         layer_dense(units = 100, activation = 'relu', input_shape = c(13)) %>%
        layer_dropout(rate=0.4)  %>%
        layer_dense(units = 50, activation = 'relu')  %>%
        layer_dropout(rate=0.2)  %>%
         layer_dense(units = 1)
model %>% compile(loss = 'mse', optimizer = 'rmsprop', metrics = 'mae') 

mymodel <- model %>%          
fit(training,trainingtarget,
             epochs = 100,
             batch_size = 32,
             validation_split = 0.2,
             verbose = 0)
```

## Prediction
```{r}
model %>% evaluate(test, testtarget)
pred <- model %>% predict(test)
mean(sqrt((testtarget-pred)^2))
```
```{r}
plot(testtarget, pred) 
```
